---
title: "A Gapminder exploration"
output: pdf_document
---
```{r package_setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

knitr::opts_chunk$set(
  fig.width = 9,   # increase width
  fig.height = 9,
  out.width = "100%"   # VERY IMPORTANT
)

library(tidyr)
library(dplyr)
library(ggplot2)

source(here::here('notebooks', 'utils.R'))
```


# Data loading
```{r}
t.features = read_csv_locale(here::here('data', 'gapminder-misc.csv'))
t.countries = read_csv_locale(here::here('data', 'gapminder-countries.csv'))

t.all = left_join(t.features, t.countries, by = 'country')

col.features = colnames(t.features)[seq(3, ncol(t.features), by = 2)]
```
Gapminder data were loaded from [Gapminder](https://www.gapminder.org/data/), avalaible (observations and forecasts) from `r length(unique(t.all$country))` countries.
Available features are:

```{r}
t.feature_ranges <- {
  t_values <- t.features %>%
    pivot_longer(
      cols = -c(country, time),
      names_to = "feature",
      values_to = "value"
    ) %>%
    filter(!grepl("_is_forecast$", feature))

  t_flags <- t.features %>%
    pivot_longer(
      cols = ends_with("_is_forecast"),
      names_to = "feature",
      values_to = "is_forecast"
    ) %>%
    mutate(
      feature = sub("_is_forecast$", "", feature),
      is_forecast = is_forecast %in% c(TRUE, "True", "TRUE", "true")
    )

  t_values %>%
    left_join(t_flags, by = c("country", "time", "feature")) %>%
    mutate(is_forecast = coalesce(is_forecast, FALSE)) %>%
    group_by(feature) %>%
    summarise(
      start_observation = min(time[!is.na(value)], na.rm = TRUE),
      start_forecast = if (any(is_forecast)) min(time[is_forecast], na.rm = TRUE) else NA_real_,
      end = max(time[!is.na(value)], na.rm = TRUE),
      .groups = "drop"
    )
}

knitr::kable(
  t.feature_ranges,
  col.names = c('feature', 'recorded from', 'forecasted from', 'until')
)
```
# Starter plots

## One line - one country

```{r}

ggplot(
  data = t.all |> filter(name == 'Switzerland'),
  mapping = aes(
    x = time,
    y = lex,
    # lty = lex_is_forecast
  )
) +
  geom_line(lwd = 1.5) +
  # theme_minimal(base_size = 16) +
  # scale_linetype_discrete(labels = c("Observation", "Forecast")) +
  labs(
    title = 'Switzerland life expectancy over time',
    subtitle = 'Gapminder publishe observations and forecasts',
    x = 'Year',
    y = 'Life expectancy (year)',
    # lty = ''
  )

```

## Mutliple countries

Goal:
 - add one color for each country.
 - the legend for colors must say: "Country"
 - write a caption to tell the story.

```{r}

ggplot(
  data = t.all |> filter(name %in% c('Switzerland', 'Ethiopia')),
  mapping = aes(
    x = time,
    y = lex,
    lty = lex_is_forecast,
  )
) +
  geom_line(lwd = 1.5) +
  theme_minimal(base_size = 16) +
  scale_linetype_discrete(labels = c("Observation", "Forecast")) +
  labs(
    title = 'Switzerland and Ethiopia life expectancy over time',
    subtitle = 'Gapminder publishe observations and forecasts',
    x = 'Year',
    y = 'Life expectancy (year)',
    lty = '',
  )

```

## Child death under 5 vs wealth inequality (Gini coefficient)

Goal:
 - snapshot in 2021
 - x/y with the title metrics
 - circle size with the population
 - color with the region (4 levels)

